Reproduction frequency and offspring survival decline in elephant seals past prime age

### 2024-10-10 paper/paperv3.qmd
Fixing an error in one of the models (singular fit); adding phenology to the detailed model outputs.

sighting_img <- sealdat %>% 
  group_by(animalID) %>% 
  summarize(yearborn = yearborn[1],
            first_obs = min(year),
            last_obs = max(year),
            .groups = "keep") %>% 
  reframe(yearborn = yearborn,
          year = yearborn:last_obs,
          age = year - yearborn,
          prerecruit = year < first_obs) %>% 
  left_join(sealdat %>% 
              group_by(animalID, year) %>% 
              summarize(observed_int = max(observed_int),
                        .groups = "drop"),
            by = c("animalID", "year")) %>% 
  mutate(
    observed = case_when(
      age == 0 ~ "Born",
      prerecruit ~ "Pre-recruit",
      observed_int == 0 ~ "Non-breeder",
      observed_int == 1 ~ "Breeder",
      is.na(observed_int) ~ "Unobserved"
    ),
    observed = factor(observed, levels = c("Born",
                                           "Pre-recruit",
                                           "Non-breeder",
                                           "Breeder",
                                           "Unobserved"))
  ) %>% 
  group_by(animalID) %>% 
  mutate(lastrepro = max(age),
         yearborn = na.omit(yearborn)[1]) %>% 
  ungroup() %>% 
  arrange(yearborn, lastrepro, animalID) %>% 
  mutate(animalID = factor(animalID, levels = unique(animalID))) %>% 
  filter(year > 1987) %>% 
  ggplot(aes(year, animalID)) +
  geom_raster(aes(fill = observed)) +
  scale_fill_viridis_d(direction = -1) +
  guides(fill = guide_legend(position = "inside", 
                             direction = "horizontal",
                             ncol = 1,
                             byrow = TRUE)) +
  labs(y ="Individual") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        legend.position = c(0.1, 0.95),
        legend.justification = c(0, 1),
        legend.background = element_blank(),
        legend.key.size = unit(4, "mm"),
        legend.text = element_text(size = 8))

@@

# Fit pup *recruitment* model (female offspring only)
pup_recr_mod <- glmer(
  # NOTE: no random effect for individual bc singular fit
  # In survival model, 93 individuals (15%) had >3 pups (male+female pups)
  # In recruit model, only 17 individuals (4%) had >3 pups (female pups only)
  pup_recr_int ~ age10 : age_cat + (1 | year_fct) + (1 | animalID),
  filter(surv_recr_data, pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

@@

# Are hypotheses supported by the model?
extract_p <- function(h, tails = 1) {
  if (tails == 1) {
    map_dbl(mod_sens, \(m) pnorm(m[[h]]["age10:age_catOld", "z value"]))
  } else if (tails == 2) {
    map_dbl(mod_sens, \(m) 2 * pnorm(m[[h]]["age10:age_catOld", "z value"],
                                     lower.tail = (m[[h]]["age10:age_catOld", "z value"] < 0)
                                     ))
  } else {
    stop("tails must be 1 or 2")
  }
}
extract_aic <- function(h) {
  map_dbl(mod_sens, paste(h, "aic", sep = "_"))
}
p_tbl <- tibble(age_cutoff = cutoff, 
                "H1 (breeding)" = extract_p("h1", tails = 2),
                "H2a (survival)" = extract_p("h2_surv", tails = 2),
                "H2b (recruitment)" = extract_p("h2_recr", tails = 2),
                "H3 (sex ratio; 2 tail)" = extract_p("h3", tails = 2))
aic_weight <- function(x) {
  rellik <- exp(-0.5 * (x - min(x)))
  rellik / sum(rellik)
}
aic_tbl <- tibble(age_cutoff = cutoff,
                  "H1 (breeding)" = extract_aic("h1"),
                  "H2a (survival)" = extract_aic("h2_surv"),
                  "H2b (recruitment)" = extract_aic("h2_recr"),
                  "H3 (sex ratio)" = extract_aic("h3")) %>% 
  mutate(across(-age_cutoff, aic_weight))

>>>>

sighting_img <- sealdat %>% 
  group_by(animalID) %>% 
  summarize(yearborn = yearborn[1],
            first_obs = min(year),
            last_obs = max(year),
            .groups = "keep") %>% 
  reframe(yearborn = yearborn,
          year = yearborn:last_obs,
          age = year - yearborn,
          prerecruit = year < first_obs) %>% 
  left_join(sealdat %>% 
              group_by(animalID, year) %>% 
              summarize(observed_int = max(observed_int),
                        .groups = "drop"),
            by = c("animalID", "year")) %>% 
  mutate(
    observed = case_when(
      age == 0 ~ "Born",
      prerecruit ~ "Pre-recruit",
      observed_int == 0 ~ "Non-breeder",
      observed_int == 1 ~ "Breeder",
      is.na(observed_int) ~ "Unobserved"
    ),
    observed = factor(observed, levels = c("Born",
                                           "Pre-recruit",
                                           "Non-breeder",
                                           "Breeder",
                                           "Unobserved"))
  ) %>% 
  group_by(animalID) %>% 
  mutate(lastrepro = max(age),
         yearborn = na.omit(yearborn)[1]) %>% 
  ungroup() %>% 
  arrange(yearborn, lastrepro, animalID) %>% 
  mutate(animalID = factor(animalID, levels = unique(animalID))) %>% 
  filter(year > 1987) %>% 
  ggplot(aes(year, animalID)) +
  geom_raster(aes(fill = observed)) +
  scale_fill_brewer(palette = "YlGnBu", direction = 1) +
  guides(fill = guide_legend(position = "inside", 
                             direction = "horizontal",
                             ncol = 1,
                             byrow = TRUE)) +
  labs(y ="Individual") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        legend.position = c(0.1, 0.95),
        legend.justification = c(0, 1),
        legend.background = element_blank(),
        legend.key.size = unit(4, "mm"),
        legend.text = element_text(size = 8))

@@

# Fit pup *recruitment* model (female offspring only)
pup_recr_mod <- glmer(
  # NOTE: no random effect for individual bc singular fit
  # In survival model, 93 individuals (15%) had >3 pups (male+female pups)
  # In recruit model, only 17 individuals (4%) had >3 pups (female pups only)
  pup_recr_int ~ age10 : age_cat + (1 | year_fct),
  filter(surv_recr_data, pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

```{r}
#| label: detailed-h4

@@

# Are hypotheses supported by the model? note to future readers...this part could have been done way easier, and you will have to trust our process that this was the path of least resistence for us at the point of final revisions!
extract_p <- function(h, tails = 1) {
  if (tails == 1) {
    map_dbl(mod_sens, \(m) pnorm(m[[h]]["age10:age_catOld", "z value"]))
  } else if (tails == 2) {
    map_dbl(mod_sens, \(m) 2 * pnorm(m[[h]]["age10:age_catOld", "z value"],
                                     lower.tail = (m[[h]]["age10:age_catOld", "z value"] < 0)
                                     ))
  } else {
    stop("tails must be 1 or 2")
  }
}
extract_aic <- function(h) {
  map_dbl(mod_sens, paste(h, "aic", sep = "_"))
}
p_tbl <- tibble(age_cutoff = cutoff, 
                "H1 (breeding)" = extract_p("h1", tails = 2),
                "H2a (survival)" = extract_p("h2_surv", tails = 2),
                "H2b (recruitment)" = extract_p("h2_recr", tails = 2),
                "H3 (sex ratio; 2 tail)" = extract_p("h3", tails = 2))
aic_weight <- function(x) {
  rellik <- exp(-0.5 * (x - min(x)))
  rellik / sum(rellik)
}
aic_tbl <- tibble(age_cutoff = cutoff,
                  "H1 (breeding)" = extract_aic("h1"),
                  "H2a (survival)" = extract_aic("h2_surv"),
                  "H2b (recruitment)" = extract_aic("h2_recr"),
                  "H3 (sex ratio)" = extract_aic("h3")) %>% 
  mutate(across(-age_cutoff, aic_weight))
